---
title: 'Week 5: Final'
author: "COVID Data Analysis"
date: "2/27/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
```{r libraries}
install.packages("tidyverse")
library(tidyverse)
library(lubridate)
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_global.csv",
                "time_series_covid19_confirmed_US.csv",
                "time_series_covid19_deaths_US.csv")
urls <- str_c(url_in, file_names)
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
US_cases <- read_csv(urls[3])
US_deaths <- read_csv(urls[4])
global_cases <- global_cases %>% pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long), names_to = "date", values_to = "cases") %>% select(-c(Lat,Long))
global_deaths <- global_deaths %>% pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long), names_to = "date", values_to = "deaths") %>% select(-c(Lat,Long))
global <- global_cases %>% full_join(global_deaths) %>% rename(Country_Region = 'Country/Region', Province_State = 'Province/State') %>% mutate(date = mdy(date))
summary(global)
global <- global %>% filter(cases > 0)
summary(global)
US_cases <- US_cases %>% pivot_longer(cols = -(UID:Combined_Key), names_to = "date", values_to = "cases") %>% select(Admin2:cases) %>% mutate(date = mdy(date)) %>% select(-c(Lat, Long_))
US_deaths <- US_deaths %>% pivot_longer(cols = -(UID:Population), names_to = "date", values_to = "deaths") %>% select(Admin2:deaths) %>% mutate(date = mdy(date)) %>% select(-c(Lat, Long_))
US <- US_cases %>% full_join(US_deaths)
summary(US)
global <- global %>% unite("Combined_Key", c(Province_State, Country_Region), sep = ", ", na.rm = TRUE, remove = FALSE)
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url) %>% select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))
global <- global %>% left_join(uid, by = c("Province_State", "Country_Region")) %>% select(-c(UID, FIPS)) %>% select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)
US_by_state <- US %>% group_by(Province_State, Country_Region, date) %>% summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% mutate(deaths_per_mill = deaths * 1000000 / Population) %>% select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()
US_totals <- US_by_state %>% group_by(Country_Region, date) %>% summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% mutate(deaths_per_mill = deaths * 1000000 / Population) %>% select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()
US_totals %>% filter(cases > 0) %>% ggplot(aes(x = date, y = cases)) + geom_line(aes(color = "cases")) + geom_point(aes(color = "cases")) + geom_line(aes(y = deaths, color = "deaths")) + geom_point(aes(y = deaths, color = "deaths")) + scale_y_log10() + theme(legend.position = "bottom", axis.tex.x = element_text(angle = 90)) + labs(title = "COVID19 in US", y = NULL)

US_by_state <- US_by_state %>% mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))
US_totals <- US_totals %>% mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))
US_state_totals <- US_by_state %>% summarize(deaths = max(deaths), cases = max(cases), population = max(Population), cases_per_thou = cases * 1000 / population, deaths_per_thou = deaths * 1000 / population) %>% filter(cases > 0, population > 0)

mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)
summary(mod)
US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))
US_tot_w_pred %>% ggplot() + geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") + geom_point(aes(x = cases_per_thou, y = pred), color = "red")
```

Import, tidy and analyze the COVID19 dataset from the Johns Hopkins github site. This is the same dataset I used in class. Feel free to repeat and reuse what I did if you want to. Be sure your project is reproducible and contains some visualization and analysis that is unique to your project. You may use the data to do any analysis that is of interest to you. You should include at least two visualizations and one model.  Be sure to identify any bias possible in the data and in your analysis.

Ideas: Predictor state per region? Predictor state for a group of states?